Reducing RES Droughts through the integration of wind and solar PV

### 2025-04-11 Article/RES_droughts_Morin_ReEn_R3.tex
Revert "Update RES_droughts_Morin_ReEn_R3.tex"
This reverts commit e5261afa93834a1c33017bc2040b79d91e3056d9.

The EU aims to generate at least 69\% of its electricity from renewable energy sources (RES) by 2030, up from 41\% in 2022~\citep{eurostat2023share}. While this transition is essential for reducing greenhouse gas emissions, it also highlights the challenge of managing the variability of weather-dependent energy sources such as wind and solar photovoltaic (PV) power. This challenge is amplified by the increasing electrification of energy sectors, which places greater demand on the power system and makes it more sensitive to meteorological conditions, both in historical~\citep{bloomfield2016} and future climates~\citep{bloomfield2021}. Periods of low renewable generation, known as \textit{Dunkelflaute} or RES droughts, pose significant risks to system adequacy and energy security, emphasising the need for a resilient energy system to meet both growing electricity demand and decarbonisation targets.

@@

The wind CF in C3SE was calculated using wind speeds at 100 metres ($u_{100}$, $v_{100}$) and a standard turbine model, the Vestas V136/3450, with a fixed hub height of 100 meters. As data on wind turbine fleet locations and specifications are difficult to obtain across Europe, C3SE assumes a homogeneous distribution of wind turbines across the ERA5 grid. While this approach does not capture the precise capacity factors reported by grid operators, it provides a well-correlated time series that effectively represents the impact of climate variability on wind power generation. The C3SE solar PV CF was also calculated for the ERA5 grid. It is derived from meteorological data, including surface solar radiation downwards ($ssrd$) and air temperature ($t2m$), using a reference solar PV plant model. This model incorporates empirical calculations for key system components such as optical losses, module efficiency, and inverters. The final CF accounts for a mix of module orientations typical for each location~\citep{saintdrenan2018solar}.

@@

The C3S~GRD dataset uses, as inputs, the actual locations of the RES farms in Ireland, and the CF from C3SE over the ERA5 grid. For each farm, the CF from the nearest grid point on the C3SE dataset was selected. A weighted average of the CF associated with each farm, using the farm's installed capacities, was used to produce the total CF time series.

@@

The ATL dataset is produced using the Atlite model. Atlite allows the user to define the wind turbine power curve and PV panel model to use when converting weather variables to wind and solar PV generation. The Atlite model takes as inputs the locations of RES farms and ERA5 weather variables: wind speed at 100 metres ($u_{100}$, $v_{100}$) for wind generation, and radiation variables ($ssr$, $ssrd$, $tisr$, and $fdir$) along with air temperature ($t2m$) for solar PV generation. The output of the Atlite model is a generation time series, which is divided by the total capacity to transform it back into CF. The selection of the wind turbine power curve and PV panel model represents the key difference between this dataset and C3S~GRD. This study identifies the most appropriate wind turbine power curve to use from the 121 power curves, each at five different levels of smoothing, made available by Renewables.ninja~\citep{staffell2016wake}, and selects the PV panel model out of the options available within Atlite.

@@

New time series were generated for both the ATL and C3S~GRD solar PV datasets, incorporating a revised distribution of installed capacity across Ireland as specified in the roadmap \textbf{[ADD REFERENCE]}. For wind power, the CF time series remains unchanged, as significant shifts in the location of wind farms are not expected. In total, twelve CF time series were analysed in this study, six for individual wind and solar PV CF (three datasets for each source) in the 91W-9PV scenario, and an additional six time series that include the combined CF for 91W-9PV and 57W-43PV scenarios across the different datasets.

@@

\begin{itemize}
	\item Wind Power - based on the actual capacity at the end of 2023
	\item Solar PV Power - based on the actual capacity at the end of 2023
	\item Combined RES / 91W-9PV - based on the actual capacity at the end of 2023
	\item Combined RES / 57W-43PV - based on the projected capacity for 2030
\end{itemize}

@@

The accuracy of the datasets used in this study was verified, before continuing to the analysis of RES droughts. For the verification process, time-varying values of installed capacity were used to account for changes in RES development over the verification period. This step allowed us to assess how well the datasets represent the production of renewable energy by comparing them against observed data. This validation step evaluates how well the datasets represent actual renewable energy production by comparing them against observed data. The overall statistical distribution of CF values for wind (2014–2023) and solar PV (2023) is presented in the violin plots in Fig.~\ref{fig:violin_plots}. These plots illustrate the density of CF values for each dataset, highlighting their differences and alignment with observations. The results indicate that ATL aligns more closely with OBS for wind, while all datasets exhibit similar distributions for solar PV.

@@

This verification for wind generation data highlights the importance of selecting a representative wind turbine power curve for the region being analysed. The ATL dataset, which uses a representative wind turbine power curve, is skilled at reproducing wind CF and RES droughts over Ireland. On the other hand, the power curve used for both C3S~GRD and C3S~NAT is not representative for Ireland, as it severely overestimates generation, underestimating the occurrence of RES droughts. This highlights a problem with using generalised datasets for analysing RES droughts: biases severely affect their ability to accurately reproduce RES drought events. The skill scores for the three datasets (Tab.~\ref{tab:wind_skill_scores}) show only a small difference in their ability to reproduce the changes in CF, as seen by their similar CC scores. However, their ability to reproduce the actual CF values is much lower than that of ATL, with RMSE scores almost four times bigger for the two C3S datasets. There is a clear bias towards an overestimation of CF, seen in the MBE values, which leads to the underestimation of RES droughts. This highlights the need to use regionally verified datasets to assess RES droughts.

@@

RES drought events identified over the 45-year period were used to calculate the return periods for different RES drought durations. A return period is the estimated average time interval between events of a specified duration (not to be confused with the frequency of their occurrence within a fixed time frame). Fig.~\ref{fig:return_periods} shows the return periods for different RES drought durations, which can be used to capture the most extreme events affecting the system. Understanding their return periods is crucial, as extreme yet rare RES droughts pose the toughest challenge to energy security by placing significant strain on the conventional backup sources necessary to maintain security of supply during these events.

@@

Under the 91W-9PV scenario (Fig.~\ref{fig:return_periods}c), the combined RES drought return periods mirror those for wind alone, reflecting the dominance of wind in the current energy mix. In contrast, the 57W-43PV scenario (Fig.~\ref{fig:return_periods}d) shows a dramatic reduction in RES drought durations, suggesting that a more balanced share of wind and solar PV capacity can substantially mitigate the frequency of prolonged RES drought events. For example, the return period for a five-day RES drought event (shown by the vertical dashed lines in Fig.~\ref{fig:return_periods}) increases from roughly six months for the 91W-9PV scenario, to four years for the 57W-43PV scenario in the ATL dataset, and from about fifteen months to around five years in the two C3S datasets. This result indicates that the complementarity between wind and solar PV plays a crucial role in reducing the occurrence of RES drought events in a diversified energy portfolio.

@@

The 91W-9PV scenario (Fig.~\ref{fig:res_droughts_seasonality}c) shows patterns similar to the ones for wind droughts (Fig.~\ref{fig:res_droughts_seasonality}a). However, in the 91W/9PV scenario, the number of hours classified as RES droughts in summer decreases slightly compared to the wind-only scenario. This reduction can be explained by the contribution of solar PV generation during the summer months in the 91W-9PV scenario, even though it constitutes only 11\% \textbf{[SHOULD THIS BE 9\%?]}of total capacity. Since the number of RES drought hours for solar PV in summer is near zero, this small contribution has a noticeable impact on reducing overall RES drought hours. In the 57W-43PV scenario (Fig.~\ref{fig:res_droughts_seasonality}d), all three datasets show a reduction in monthly RES drought frequency. Annual reductions in median RES drought frequency are observed across the datasets, dropping from 14\% to 5\% for ATL, from 8\% to 3\% for C3S~GRD, and from 9\% to 4\% for C3S~NAT. The balanced mix of wind and solar PV power in this scenario reduces the seasonal signal overall and significantly decreases the percentage of RES drought hours in the summer.

>>>>

The EU aims to generate at least 69\% of its electricity from renewable energy sources (RES) by 2030, up from 41\% in 2022~\citep{eurostat2023share}. While this transition is essential for reducing greenhouse gas emissions, it also highlights the challenge of managing the variability of weather-dependent energy sources such as wind and solar photovoltaic (PV) power. This challenge is amplified by the increasing electrification of energy sectors, which places greater demand on the power system and makes it more sensitive to meteorological conditions~\citep{staffell2018demand}. Periods of low renewable generation, known as \textit{Dunkelflaute} or RES droughts, pose significant risks to system adequacy and energy security, emphasising the need for a resilient energy system to meet both growing electricity demand and decarbonisation targets.

@@

The wind CF in C3SE was calculated using wind speeds at 100 m ($u_{100}$, $v_{100}$) and a standard wind turbine model, the Vestas V136/3450, with a fixed hub height of 100 m. As data on wind turbine fleet locations and specifications are difficult to obtain across Europe, C3SE assumes a homogeneous distribution of wind turbines across the ERA5 grid. While this approach does not capture the precise capacity factors reported by grid operators, it provides a well-correlated time series that effectively represents the impact of climate variability on wind power generation. The C3SE solar PV CF was also calculated for the ERA5 grid. It is derived from meteorological data, including surface solar radiation downwards ($ssrd$) and air temperature ($t2m$), using a reference solar PV plant model. This model incorporates empirical calculations for key system components such as optical losses, module efficiency, and inverters. The final CF accounts for a mix of module orientations typical for each location~\citep{saintdrenan2018solar}.

@@

The C3S~GRD dataset uses, as inputs, the actual locations of the RES farms in Ireland, and the CF from C3SE over the ERA5 grid. For each farm, the CF from the nearest grid point on the C3SE dataset was selected. A weighted average of the CF associated with each farm, using the farm's installed capacities, was used to produce the combined CF time series.

@@

The ATL dataset is produced using the Atlite model. Atlite allows the user to define the wind turbine power curve and PV panel model to use when converting weather variables to wind and solar PV generation. The Atlite model takes as inputs the locations of RES farms and ERA5 weather variables: wind speed at 100 m ($u_{100}$, $v_{100}$) for wind generation, and radiation variables ($ssr$, $ssrd$, $tisr$, and $fdir$) along with air temperature ($t2m$) for solar PV generation. The output of the Atlite model is a generation time series, which is divided by the total capacity to transform it back into a CF. The selection of the wind turbine power curve and PV panel model represents the key difference between this dataset and C3S~GRD. This study identifies the most appropriate wind turbine power curve to use from the 121 power curves, each at five different levels of smoothing, made available by Renewables.ninja~\citep{staffell2016wake}, and selects the PV panel model out of the options available within Atlite.

@@

New time series were generated for both the ATL and C3S~GRD solar PV datasets, incorporating a revised distribution of installed capacity across Ireland as specified in the roadmap~\citep{eirgrid2023future}. For wind power, the CF time series remains unchanged, as significant shifts in the location of wind farms are not expected. In total, twelve CF time series were analysed in this study, six for individual wind and solar PV CF (three datasets for each source) in the 91W-9PV scenario, and an additional six time series that include the combined CF for 91W-9PV and 57W-43PV scenarios across the different datasets.

@@

\begin{itemize}
	\item Wind Power - based on actual capacity at the end of 2023
	\item Solar PV Power - based on actual capacity at the end of 2023
	\item Combined RES / 91W-9PV - based on actual capacity at the end of 2023
	\item Combined RES / 57W-43PV - based on projected capacity for 2030
\end{itemize}

@@

The accuracy of the datasets used in this study was verified, before continuing to the analysis of RES droughts. For the verification process, time-varying values of installed capacity were used to account for changes in RES development over the verification period. This step allowed us to assess how well the datasets represent renewable generation profiles by comparing them against observed data. This validation step evaluates how well the datasets represent actual renewable energy production by comparing them against observed data. The overall statistical distribution of CF values for wind (2014–2023) and solar PV (2023) is presented in the violin plots in Fig.~\ref{fig:violin_plots}. These plots illustrate the density of CF values for each dataset, highlighting their differences and alignment with observations. The results indicate that ATL aligns more closely with observations for wind, while all datasets exhibit similar distributions for solar PV.

@@

Verification of wind generation data highlights the importance of selecting a representative wind turbine power curve for the region being analysed. The ATL dataset, which uses a representative wind turbine power curve, is skilled at reproducing wind CF and RES droughts across Ireland. On the other hand, the power curve used for both C3S~GRD and C3S~NAT is not representative for Ireland, as it severely overestimates generation, underestimating the occurrence of RES droughts. This highlights a problem with using generalised datasets for analysing RES droughts: biases severely affect their ability to accurately reproduce RES drought events. The skill scores for the three datasets (Table~\ref{tab:wind_skill_scores}) show only a small difference in their ability to reproduce the changes in CF, as seen by their similar CC scores. However, their ability to reproduce the actual CF values is much lower than that for ATL, with RMSE scores almost four times higher than for the two C3S datasets. There is a clear bias towards an overestimation of CF, seen in the MBE values, which leads to an underestimation of RES droughts. This highlights the need to use regionally verified datasets to assess RES droughts.

@@

RES drought events identified over the 45-year period were used to calculate the return periods for different RES drought durations. A return period is the estimated average time interval between events of a specified duration (not to be confused with the frequency of their occurrence within a fixed time frame). Fig.~\ref{fig:return_periods} shows the return periods for different RES drought durations, which can be used to capture the most extreme events affecting the system. Understanding their return periods is crucial, as extreme but rare RES droughts pose the toughest challenge to energy security by placing significant strain on the conventional backup sources necessary to maintain security of supply during these events.

@@

Under the 91W-9PV scenario (Fig.~\ref{fig:return_periods}c), the combined RES drought return periods mirror those for wind alone, reflecting the dominance of wind in the current energy mix. In contrast, the 57W-43PV scenario (Fig.~\ref{fig:return_periods}d) shows a dramatic reduction in RES drought durations, suggesting that a more balanced share of wind and solar PV capacity can substantially mitigate the frequency of prolonged RES drought events. For example, the return period for a five-day RES drought event (shown by the vertical dashed lines in Fig.~\ref{fig:return_periods}) increases from roughly six months for the 91W-9PV scenario, to four years for the 57W-43PV scenario for the ATL dataset, and from about fifteen months to around five years for the two C3S datasets. This result indicates that the complementarity between wind and solar PV plays a crucial role in reducing the occurrence of RES drought events in a diversified energy portfolio.

@@

The 91W-9PV scenario (Fig.~\ref{fig:res_droughts_seasonality}c) shows patterns similar to the ones for wind droughts (Fig.~\ref{fig:res_droughts_seasonality}a). However, in the 91W/9PV scenario, the number of hours classified as RES droughts in summer decreases slightly compared to the wind-only scenario. This reduction can be explained by the contribution of solar PV generation during the summer months in the 91W-9PV scenario, even though it constitutes only 9\% of total capacity. Since the number of RES drought hours for solar PV in summer is near zero, this small contribution has a noticeable impact on reducing overall RES drought hours. In the 57W-43PV scenario (Fig.~\ref{fig:res_droughts_seasonality}d), all three datasets show a reduction in monthly RES drought frequency. Annual reductions in median RES drought frequency are observed across the datasets, dropping from 14\% to 5\% for ATL, from 8\% to 3\% for C3S~GRD, and from 9\% to 4\% for C3S~NAT. The balanced mix of wind and solar PV power in this scenario reduces the seasonal signal overall and significantly decreases the percentage of RES drought hours in the summer.
